Persistence in functional topology: local-connectivity, q-tameness and a correction to a theorem of Morse

### 2021-04-09 connectivity.tex
Add section by section to intro
Improve section, connectivity: remove Bredon.

\section{Local connectivity and q-tameness}

@@

Local connectivity assumptions have a long history of being used to ensure that different homology theories agree and to ensure that homology is finite dimensional, see \cite{MR0007094} for early examples.
For us, the following is relevant.
Recall that a space is said to be \textit{paracompact} if every open cover has a locally finite open refinement, i.e., every point has a neighborhood intersecting finitely many sets in the refinement.

\begin{prop}[{\cite{MR105677, MR1481706}}] \label{prop:cech_sing_hom_hlc}
	On the category of paracompact Hausdorff $\HLC$ spaces, \v{C}ech and singular homology with arbitrary coefficients coincide.
\end{prop}

For more results of this type, we refer to \cite{MR1481706}. There, the above proposition \cite[Corollary VI.12.6]{MR1481706} and many similar ones are proven using sheaf theoretic methods.
For example, Bredon also discusses cohomology local connectedness \cite[Section II.17]{MR1481706} and various other comparison results, e.g.\@ between singular and Borel--Moore homology \cite[Corollary V.12.15]{MR1481706} in the presence of homology local connectedness, as well as examples such as spaces that are sheaf cohomology locally connected but not homology locally connected \cite[Example II.17.12]{MR1481706}.

For sublevel filtrations we obtain the following persistence version of Proposition~\ref{prop:cech_sing_hom_hlc} obtained by applying it pointwise.

\begin{cor}\label{cor:cech_sing_persistent_iso}
	Let $(X, f)$ be a sublevel filtration. If each sublevel set is paracompact and $\HLC$, then its singular and \v{C}ech persistence modules coincide.
\end{cor}

We are interested in statements that bound the dimension of homology with field coefficients.
In the non-filtered context one has the following result.

\begin{prop} \label{prop:fin_dim_sing_hom}
	Let $X$ be a compact Hausdorff $\HLC$ space.
	Then the singular homology of $X$ with field coefficients is finite-dimensional in every degree.
\end{prop}

\begin{proof}
	By the universal coefficient theorem for homology, trivial singular homology with integer coefficients also implies trivial singular homology for any other coefficient group. In particular, this means that $X$ being $HLC^{\infty}$ also implies that $X$ is $HLC^{\infty}_{\mathbb{F}}$. By the universal coefficient theorem for cohomology, the singular cohomology with coefficients in $\mathbb{F}$ is naturally isomorphic to the dual space of the singular homology with coefficients in $\mathbb{F}$. This means that being $HLC^{\infty}_{\mathbb{F}}$ implies being locally connected with respect to singular cohomology in all degrees with coefficients in $\mathbb{F}$. Since we assume $X$ to be $HLC^{\infty}$, \cite[Theorem III.1.1]{MR1481706} now implies that $X$ is locally connected with respect to sheaf cohomology with coefficients in the locally constant sheaf with value $\mathbb{F}$. \cite[Corollary II.17.7]{MR1481706} implies that the sheaf cohomology of the whole space with coefficients in this sheaf is finite-dimensional, so again by \cite[Theorem III.1.1]{MR1481706} the same holds for singular cohomology with coefficients in $\mathbb{F}$. Applying the universal coefficient theorem once more proves the claim.
\end{proof}

Again, there is a persistent version for sublevel filtrations.

\begin{cor}
	Let $(X, f)$ be a sublevel filtration. If each sublevel set is compact and $\HLC$, then its singular persistence module is p.f.d. in every degree, in particular q-tame.
\end{cor}

Having each space in a sublevel filtration being $\HLC$ is in many cases too restrictive.
We would like a relative condition for sublevel filtrations to be sufficient to ensure the q-tameness of the associated persistence module.

\begin{defi} \label{defi:local_connectedness_filtrations}
	For a fix homology theory $\H$. A sublevel filtration $(X,f)$ is said to be $\HLC$ if for any $\epsilon > 0$, $x \in X$, and $V$ an open neighborhood of $x$, there is a $\delta \in (0, \epsilon)$ and a neighborhood $U$ of $x$ such that $U \subseteq V$ and
	\begin{equation*}
	X_{\leq f(x) + \delta + c} \cap U \to X_{f(x) + \epsilon + c} \cap V
	\end{equation*}
	is $\HT$ for every $c \geq 0$.
	We also consider a weaker version, referred to as \textit{weakly} $\HLC$, defined as above with $c = 0$.
\end{defi}

We remark that if $\H$ is singular homology, both conditions of $\HLC$ for a Morse filtration $(M,f)$ are respectively equivalent to Morse's definitions of locally $f$-connectedness when $M$ is a metric space.
Please consult the historical recount presented in Section~\ref{sec:historical}.

We will now show that the weak notion does not imply q-tameness in general.
To do so let us consider the \textit{$d$-dimensional Hawaiian earring}
\begin{equation*}
\HE = \bigcup_{n\in\mathbb{N}}\left\{(x_0,\dots,x_d)\in\R^{d+1} \ \middle | \ \left(x_0-\frac{1}{n}\right)^2+x_1^2+\dots+x_d^2=\left(\frac{1}{n}\right)^2\right\}.
\end{equation*}
\todo{credit to Milnor}

@@

\begin{prop}
	The assumptions of Theorem~\ref{thm:counterexample} are satisfied by singular homology with any coefficients and \v{Cech} homology with field coefficients.
\end{prop}

>>>>

\section{Local connectivity and q-tameness} \label{s:connectivity}

@@

Local connectivity assumptions have a long history of being used to ensure that different homology theories agree and to ensure that homology is finite dimensional.
We mention the following examples.

Recall that a space is said to be \textit{paracompact} if every open cover has a locally finite open refinement, i.e., every point has a neighborhood intersecting finitely many sets in the refinement.
\begin{prop}[\cite{MR105677}] \label{prop:cech_sing_hom_hlc}
	\v{C}ech and singular homology with arbitrary coefficients coincide on the category of paracompact Hausdorff spaces which are $\HLC$ with respect to integral singular homology.
\end{prop}

\begin{prop} \label{prop:fin_dim_sing_hom}
	Let $X$ be a compact Hausdorff space.
	If $X$ is $\HLC$ with respect to integral singular homology, then the singular homology of $X$ with arbitrary coefficients is finite-dimensional in every degree.
\end{prop}

We will deduce the previous finiteness result from our main contribution, Theorem~\ref{t:strong local connectenss implies q-tameness}.
To motivate the statement of our theorem let us record a direct consequence of the previous proposition.

\begin{prop}
	Let $(X, f)$ be a sublevel filtration. If each sublevel set is compact and $\HLC$ with respect to integral singular homology, then its singular persistence module is p.f.d. in every degree, in particular q-tame.
\end{prop}

Having each sublevel set being $\HLC$ is in many cases too restrictive.
We would like a relative condition for sublevel filtrations to be sufficient to ensure the q-tameness of the associated persistence module.
The following definitions are motivated by Morse's work, please consult Section~\ref{s:historical} for their relation to his original statements.

\begin{defi} \label{defi:local_connectedness_filtrations}
	For a fix homology theory $\H$. A sublevel filtration $(X,f)$ is said to be $\HLC$ if for any $\epsilon > 0$, $x \in X$, and $V$ an open neighborhood of $x$, there is a $\delta \in (0, \epsilon)$ and a neighborhood $U$ of $x$ such that $U \subseteq V$ and
	\begin{equation*}
	X_{\leq f(x) + \delta + c} \cap U \to X_{\leq f(x) + \epsilon + c} \cap V
	\end{equation*}
	is $\HT$ for every $c \geq 0$.
	We also consider a weaker version, referred to as \textit{weakly} $\HLC$, defined as above with $c = 0$.
\end{defi}

We will now show that the weak notion does not imply q-tameness in general.
To do so let us consider the \textit{$d$-dimensional Hawaiian earring}
\begin{equation*}
\HE = \bigcup_{n\in\mathbb{N}}\left\{(x_0,\dots,x_d)\in\R^{d+1} \ \middle | \ \left(x_0-\frac{1}{n}\right)^2+x_1^2+\dots+x_d^2=\left(\frac{1}{n}\right)^2\right\}.
\end{equation*}

@@

\begin{prop}
	The assumptions of Theorem~\ref{thm:counterexample} are satisfied by singular homology with arbitrary coefficients and \v{Cech} homology with field coefficients.
\end{prop}

### 2021-04-09 introduction.tex
Improve section, connectivity: remove Bredon.

The interplay between the critical set of a function and the topology of its domain is a cornerstone of modern mathematics.
Now a days, when thinking about Morse's work a smooth function on a closed differential manifold is possibly our first association, but it need not be our last.
Morse Theory in this \textit{finite} context was masterfully presented in Milnor's famous book on the subject \cite{MR0163331}; where he also move beyond it to give a new proof of Bott's periodicity, applying Morse theory to the energy functional of paths in a Riemannian manifold.
Another example of the importance of Morse's insights in infinite settings is Floer's work on the Arnold conjecture.
Morse himself worked in a significantly less constraint context, publishing in the 1930s a series of papers \cite{Morse.1937, Morse.1938, Morse.1940, MR9102} in which he established the key results of Morse theory in the broad context of semi-continuous functionals on metric spaces.
He name this body of work \textit{functional topology}.
A motivating problem for him came from the study of minimal surfaces.
In 1931, Jesse Douglas solved Plateau’s Problem \cite{Douglas.1931} for which he received the fields medal in 1936.
The problem amounts to finding for a given admissible boundary curve $\Gamma$ an area-minimizing embedded surface $\Sigma$ with $\partial \Sigma = \Gamma$.
In \cite{Morse.1939, Morse.1941}, Morse and Tompkins used functional topology to give conditions for the existence of unstable minimal surfaces, i.e., critical points of the area functional that do not locally minimize area.

Some of Morse's ideas from functional topology, and in particular from \cite{Morse.1940}, have been rediscovered in topological data analysis.
Morse studied the topology of a sublevel filtration and he introduced invariants that encode the same information as the modern persistent homology groups.
Even though Morse did not have a structure theory of persistence modules via barcodes, he emphasized many points that are crucial for this point of view:
The use of homology with field coefficients and the restriction to functionals giving rise to persistence modules with interval decompositions.

@@

An overview of this article follows.
In Section~\ref{s:homology} we review different constructions of ordinary homology theory including: \v{C}ech, Vietoris and singular.
In section~\ref{s:persistence} we recall the basics of persistence theory where, for us, the persistence homology of a sublevel set filtration is the key example.
Section~\ref{s:connectivity} is the core of this work.
We define two natural notions of local-connectivity for a sublevel set filtration and show under what circumstances they imply the associated persistence module is q-tame.
We close with a historical overview of Morse's functional topology in Section~\ref{s:homology}.

>>>>

The interplay between the critical set of a function and the topology of its domain is a cornerstone of modern mathematics.
Now a days, when thinking about the pioneering work of Morse in this interplay, a smooth function on a closed differential manifold is possibly our first association but, as we describe next, it should not be our last.
Morse Theory in the \textit{finite} context was masterfully presented in Milnor's famous book on the subject \cite{MR0163331}; where he also move beyond it to give a new proof of Bott's periodicity, applying Morse theory to the energy functional of paths in a Riemannian manifold.
Another example of the importance of Morse's insights in infinite settings is Floer's work on the Arnold conjecture.
Morse himself worked in a significantly less constraint context, publishing in the 1930s a series of papers \cite{Morse.1937, Morse.1938, Morse.1940, MR9102} in which he established the key results of Morse theory in the broad context of semi-continuous functionals on metric spaces.
He name this body of work \textit{functional topology}.
A motivating problem for him came from the study of minimal surfaces.
In 1931, Jesse Douglas solved Plateau’s Problem \cite{Douglas.1931} for which he received the fields medal in 1936.
The problem amounts to finding for a given admissible boundary curve $\Gamma$ an area-minimizing embedded surface $\Sigma$ with $\partial \Sigma = \Gamma$.
In \cite{Morse.1939, Morse.1941}, Morse and Tompkins used functional topology to give conditions for the existence of unstable minimal surfaces, i.e., critical points of the area functional that do not locally minimize area.

Some of Morse's ideas from functional topology, and in particular from \cite{Morse.1940}, have been rediscovered in topological data analysis.
Morse studied the topology of a sublevel set filtration introducing invariants that encode the same information as the modern persistent homology groups.
Even though Morse did not have a structure theory of persistence modules via barcodes, he emphasized many points that are crucial from point of view:
The use of homology with field coefficients and the restriction to functionals giving rise to persistence modules with interval decompositions.

@@

An overview of this article follows.
In Section~\ref{s:homology} we review different constructions of ordinary homology theory including: \v{C}ech, Vietoris and singular.
In section~\ref{s:persistence} we recall the basics of persistence theory where, for us, the persistence homology of a sublevel set filtration is the key example.
Section~\ref{s:connectivity} is the core of this work.
We define two natural notions of local-connectivity for a sublevel set filtration and show under what circumstances they imply the associated persistence module is q-tame.
We close with a historical overview of Morse's functional topology in Section~\ref{s:historical}.
